---
title: 'STAT 420: Homework 6'
author: "Fall 2016, Unger"
date: 'Due: Monday, October 24 by 11:50 PM CT'
output:
  html_document:
    theme: readable
    toc: yes
---

# Assignment

## Exercise 1 (EPA Emissions Data)

For this exercise we will use the data stored in [`epa2015.csv`](epa2015.csv). It contains detailed descriptions of 4,411 vehicles manufactured in 2015 used for fuel economy testing [as performed by the Environment Protection Agency]( https://www3.epa.gov/otaq/tcldata.htm). The variables in the dataset are:  
 
- `Make` - manufacturer
- `Model` - model of vehicle
- `ID` - manufacturer defined vehicle identification number within EPA's computer system (not a VIN number)
- `disp` - cubic inch displacement of test vehicle
- `type` - car, truck, or both (for vehicles that meet specifications of both car and truck, like smaller SUVs or crossovers)
- `horse` - rated horsepower, in foot-pounds per second
- `cyl` - number of cylinders
- `lockup` - vehicle has transmission lockup; N or Y
- `drive` - drivetrain system code
    - A = All wheel drive
    - F = Front wheel drive
    - P = Part-time 4-wheel drive
    - R = Rear wheel drive
    - 4 = 4-wheel drive
- `weight` - test weight, in pounds
- `axleratio` - axle ratio
- `nvratio` - n/v ratio (engine speed versus vehicle speed at 50 mph)
- `THC` - total hydrocarbons, in grams per mile (g/mi)
- `CO` - Carbon monoxide (a regulated pollutant), in g/mi
- `CO2` - Carbon dioxide (the primary byproduct of all fossil fuel combustion), in g/mi
- `mpg` - fuel economy, in miles per gallon

We will attempt to model `CO2` using both `horse` and `type`. In practice we would use many more predictors, but limiting ourselves to these two, one numeric and one factor, will allow us to create a number of plots.

**(a)** Load the data, and check its structure using `str()`. Verify that `type` is a factor, if not, coerce it to be a factor. 

```{r}
epa2015 <- read.csv("~/Downloads/epa2015.csv", header=TRUE)

str(epa2015)
is.factor(epa2015$type)


```

**(b)** Make a scatterplot of `CO2` versus `horse`. Use a different color point for each vehicle `type`. Which color is which `type`?

```{r}
epa2015 <- read.csv("~/Downloads/epa2015.csv", header=TRUE)
plot(epa2015$CO2 ~ epa2015$horse, data = epa2015, col = epa2015$type)
#From this plot I can see that red is for type car, black is for both, and green is for the trucks.

```

**(c)** Fit a SLR model with `CO2` as the response and only `horse` as the predictor. Recreate your plot and add the fitted regression line. Comment on how well this line models the data. Give an estimate for the average change in `CO2` for a one foot-pound per second increase in `horse` for a vehicle of type `truck`. Give a 95% prediction interval using this model for the `CO2` of a Subaru Impreza Wagon which is a vehicle with 148 horsepower and is considered type `Both`. (Interestingly, the dataset gives the wrong drivetrain for most Subarus in this dataset, as they are almost all listed as `F`, when they are in fact All-Wheel-Drive.)

```{r}
epa2015 <- read.csv("~/Downloads/epa2015.csv", header=TRUE)

##Fit a SLR model with `CO2` as the response and only `horse` as the predictor. Recreate your plot and add the fitted regression line. Comment on how well this line models the data.
CO2_slr = lm(CO2 ~ horse, data=epa2015)
plot(CO2 ~ horse, data = epa2015, col = type)
abline(CO2_slr, col = "darkorange")
#This line seems to model the data fairly well.
```
```{r}
##Give an estimate for the average change in `CO2` for a one foot-pound per second increase in `horse` for a vehicle of type `truck`.
coef(CO2_slr)
```
```{r}
##Give a 95% prediction interval using this model for the `CO2` of a Subaru Impreza Wagon which is a vehicle with 148 horsepower and is considered type `Both`.
```
```{r}
subaru = data.frame(horse=148,type="Both")
predict(CO2_slr, subaru, interval = "prediction", level = 0.95)
```

**(d)** Fit an additive multiple regression model with `CO2` as the response and `horse` and `type` as the predictors. Recreate your plot and add the fitted regression "lines" with the same colors as their respective points. Comment on how well these lines models the data. Give an estimate for the average change in `CO2` for a one foot-pound per second increase in `horse` for a vehicle of type `truck`. Give a 95% prediction interval using this model for the `CO2` of a Subaru Impreza Wagon which is a vehicle with 148 horsepower and is considered type `Both`.

```{r}
##Fit an additive multiple regression model with `CO2` as the response and `horse` and `type` as the predictors. Recreate your plot and add the fitted regression "lines" with the same colors as their
CO2_mlr = lm(CO2 ~ horse + type, data = epa2015)
plot(CO2 ~ horse, data = epa2015, col = type)
abline(CO2_mlr, lwr = 3, col = "orange")
#Looks like a good model.
```
```{r}
##Give an estimate for the average change in `CO2` for a one foot-pound per second increase in `horse` for a vehicle of type `truck`.
coef(CO2_mlr)[2]
```
```{r}
##Give a 95% prediction interval using this model for the `CO2` of a Subaru Impreza Wagon which is a vehicle with 148 horsepower and is considered type `Both`
subaru = data.frame(horse=148,type="Both")
predict(CO2_mlr, subaru, interval = "prediction", level = 0.95)
```

**(e)** Fit an interaction multiple regression model with `CO2` as the response and `horse` and `type` as the predictors. Recreate your plot and add the fitted regression "lines" with the same colors as their respective points. Comment on how well these lines models the data. Give an estimate for the average change in `CO2` for a one foot-pound per second increase in `horse` for a vehicle of type `truck`. Give a 95% prediction interval using this model for the `CO2` of a Subaru Impreza Wagon which is a vehicle with 148 horsepower and is considered type `Both`.

```{r}
##Fit an interaction multiple regression model with `CO2` as the response and `horse` and `type` as the predictors. Recreate your plot and add the fitted regression "lines" with the same colors as their respective points.
CO2_int = lm(CO2 ~ horse*type, data=epa2015)
type_both = coef(CO2_int)[1]
type_car = coef(CO2_int)[1] + coef(CO2_int)[3]
type_truck = coef(CO2_int)[1] + coef(CO2_int)[4]
type_slope = coef(CO2_int)[2] 

plot(CO2 ~ horse, data = epa2015, col = type)
abline(type_both, type_slope, col = 1, lty = 1, lwd = 3)
abline(type_car, type_slope, col = 2, lty = 2, lwd = 3)
abline(type_truck, type_slope, col = 3, lty = 3, lwd = 3)
#These lines look like good fits
```
```{r}
##Give an estimate for the average change in `CO2` for a one foot-pound per second increase in `horse` for a vehicle of type `truck`.
est = coef(CO2_int)[2] + coef(CO2_int)[6]
est
```
```{r}
##Give a 95% prediction interval using this model for the `CO2` of a Subaru Impreza Wagon which is a vehicle with 148 horsepower and is considered type `Both`. (Same as above problems)
subaru = data.frame(horse=148,type="Both")
predict(CO2_int,subaru,interval = "prediction", level=0.95)
```

**(f)** You will perform $F$-tests later in the exercise, but for now, based solely on the three previous plots, which model is preferred: SLR, additive, or interaction?
```{r}
##The interaction model is what is preffered. This is because it better models the amount in g of co2 of the cars by looking at their horsepower and their type together.
```

**(g)** Use an ANOVA $F$-test to compare the SLR and additive models. Based on this test and a significance level of $\alpha = 0.01$, which model is preferred?
```{r}
anova(CO2_slr, CO2_mlr)

##Based on the F-stat is 203.31 and the very low p-value (< 2.2e-16) and the significance level of 0.01 we reject the null hypothesis and choose the additive model.
```

**(h)** Use an ANOVA $F$-test to compare the additive and interaction models. Based on this test and a significance level of $\alpha = 0.01$, which model is preferred?
```{r}
anova(CO2_mlr,CO2_int)

##Based on the F-stat of 11.092 and low p-value of 1.567e-05 and the significance level of 0.01 we reject the null, and choose interaction.

```
## Exercise 2 (Hospital SUPPORT Data)

For this exercise we will use the data stored in [`hospital.csv`](hospital.csv). It contains a random sample of 580 seriously ill hospitalized patients from a famous study called "SUPPORT" (Study to Understand Prognoses Preferences Outcomes and Risks of Treatment). As the name suggests, the purpose of the study was to determine what factors affected or predicted outcomes, such as how long a patient remained in the hospital. The variables in the dataset are:  
 
- `Days` - Days to death or hospital discharge
- `Age` - Age on day of hospital admission
- `Sex` - female or male
- `Comorbidity` - Patient diagnosed with more than one chronic disease
- `Charges` - Hospital charges, in dollars
- `EdYears` - Years of education
- `Education` - Education level; high or low
- `Income` - Income level; high or low
- `Care` - Level of care required; high or low
- `Race` - non-white or white
- `Pressure` - Blood pressure, in mmHg
- `WhiteBlood` - White blood cell count, in gm/dL
- `Rate` - Heart rate, in bpm

For this exercise, we will use `Charges`, `Pressure`, `Care`, and `Race` to model `Days`.

**(a)** Load the data, and check its structure using `str()`. Verify that `Care` and `Race` are factors, if not, coerce them to be factors. What are the levels of `Care` and `Race`.
```{r}
hospital <- read.csv("~/Downloads/hospital.csv")

str(hospital)

is.factor(hospital$Care)
is.factor(hospital$Race)
#levels
levels(hospital$Care)
levels(hospital$Race)

```

**(b)** Fit an additive multiple regression model with `Days` as the response using `Charges`, `Pressure`, `Care`, and `Race` as predictors. What does `R` choose as the reference level for `Care` and `Race`?
```{r}
mult = lm(Days ~ Charges+Pressure+Care+Race, data=hospital)
summary(mult)
#High for Care and Non-White for Race
```

**(c)** Fit a multiple regression model with `Days` as the response. Use the main effects of `Charges`, `Pressure`, `Care`, and `Race`, as well as the interaction of `Care` with each of the numeric predictors as predictors. (That is, the interaction of `Care` with `Charges` and the interaction of `Care` with `Pressure`.) Use a statistical test to compare this model to the additive model using a significance level of $\alpha = 0.01$. Which do you prefer?
```{r}
mult_int_reg = lm(Days ~Charges+Pressure+Care+Race+Care*Charges+Care*Pressure, data=hospital)
anova(mult, mult_int_reg)
#Based on the F-statof 4.8692 and the very low p-value of 0.008001: which is less than the significance level of 0.01 we reject the null hypothesis, thus choosing the interaction model which in is mult.
```

**(d)** Fit a multiple regression model with `Days` as the response. Use the predictors from the model in **(c)** as well as the interaction of `Race` with each of the numeric predictors. (That is, the interaction of `Race` with `Charges` and the interaction of `Race` with `Pressure`.) Use a statistical test to compare this model to the additive model using a significance level of $\alpha = 0.01$. Which do you prefer?
```{r}
mult_race = lm(Days ~Charges+Pressure+Care+Race+Care*Charges+Care*Pressure+Race*Charges+Race*Pressure, data=hospital)
anova(mult, mult_race)
#Based on the F-stat of 7.6014 and p-value of 5.699e-06 which is less than the significance level of 0.01 we reject the null hypothesis. Therefore choosing model with the interaction of Race with Charges and Pressure and interaction of Care with Charges and Pressure or mult_race.. 
```

**(e)** Using the model in **(d)** give an estimate of the change in average `Days` for a one-unit increase in `Pressure` for a `"white"` patient that required a high level of care.
```{r}
coef(mult_race)[3]+coef(mult_race)[8]
```

**(f)** Find a model using the four predictors that we have been considering that is more flexible than the model in **(d)** which is also statistical significant as compared to the model in **(d)** at a significance level of $\alpha = 0.01$.
```{r}
model = lm(Days~ Charges*Pressure*Care*Race, data= hospital)
anova(mult_race, model)

#As we can see, this interaction is much more flexible. The pvalue (1.969e-14) is less than alpha so we reject the null (the model from d), and choose model as the more flexible one.
```

## Exercise 3 (Fish Data)

For this exercise we will use the data stored in [`fish.csv`](fish.csv). It contains data for 158 fish of 7 different species all gathered from the same lake in one season. The variables in the dataset are:  
 
- `Species` - Common Name (*Latin name*)
    + 1 = Bream (*Abramis brama*)
    + 2 = Whitewish (*Leusiscus idus*)
    + 3 = Roach (*Leuciscus rutilus*)
    + 4 = <None> (*Abramis bjrkna*)
    + 5 = Smelt (*Osmerus eperlanus*)
    + 6 = Pike (*Esox Lucius*)
    + 7 = Perch (*Perca fluviatilis*)
- `Weight` - Weight of the fish, in grams
- `Length1` - Length from the nose to the beginning of the tail, in cm
- `Length2` - Length from the nose to the notch of the tail, in cm
- `Length3` - Length from the nose to the end of the tail, in cm
- `HeightPct` - Maximal height as % of Length3
- `WidthPct` - Maximal width as % of Length3
- `Sex` - 0 = female, 1 = male

We will attempt to predict `Weight` using `Length1`, `HeightPct`, and `WidthPct`.

**(a)** Use `R` to fit the model

\[
Y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_3 + \beta_4 x_1 x_2 + \beta_5 x_1 x_3 + \beta_6 x_2 x_3 + \beta_7 x_1 x_2 x_3 + \epsilon,
\]

where

- $Y$ is `Weight`,
- $x_1$ is `Length1`,
- $x_2$ is `HeightPct`,
- $x_3$ is `WidthPct`.

Report the estimated coefficients of the model.

```{r}
fish <- read.csv("~/Downloads/fish.csv", header=TRUE)
fish_model_big = lm(Weight ~ Length1*HeightPct*WidthPct, data=fish)
coef(fish_model_big)
```

**(b)** Consider fitting a smaller model in `R`.

```{r}
fish_small = lm(Weight ~ Length1 + HeightPct * WidthPct, data = fish)
```

Use a statistical test to compare this model with the previous. Report the following:

- The null and alternative hypotheses in terms of the model given in **(a)**.
- The value of the test statistic.
- The p-value of the test.
- A statistical decision using a significance level of $\alpha = 0.05$.
- Which model you prefer.

#Null hypothesis:
\[\beta_4 = \beta_5  = \beta_7 = 0  
\]
#Alternative hypothesis
#Choose one from 
\[\beta_4, \beta_5, \beta_7 
\]
# does not equal 0

```{r}
anova(fish_small,fish_model_big)

#Test-stat is 16.367 and the p-value 2.972e-09 and lower than the significance value of 0.05, so we reject the null hypothesis and prefer part a's model
```
**(c)** Give an expression based on the model in **(a)** for the true change in average weight for a one cm increase in `Length1` for a fish with a `HeightPct` of 20 and a `WidthPct` of 10.
```{r}
coef(fish_model_big)[2]+20*coef(fish_model_big)[5]+10*coef(fish_model_big)[6]+20*10*coef(fish_model_big)[8]
```
**(d)** Give an expression based on the smaller model in **(b)** for the true change in average weight for a one cm increase in `Length1` for a fish with a `HeightPct` of 20 and a `WidthPct` of 10.
```{r}
coef(fish_small)[2]

```



