---
title: 'STAT 420: Homework 10'
author: "Fall 2016, sirasao2 // Rahul Sirasao"
date: 'Due: Wednesday, December 7 by 11:50 PM CT'
output:
  html_document:
    theme: readable
    toc: yes
---

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80)
```

# Assignment

## Exercise 1 (`prostate` Data)

Using the `prostate` dataset from the `faraway` package, fit a model with `lpsa` as the response and the other variables as predictors. For this exercise only consider first order predictors.

```{r}
library(leaps)
library(faraway)
data("prostate")
```

**(a)** Find the model with the **best** AIC. Report the predictors that are used in the resulting model.

```{r}
mod_1 <- lm(lpsa ~., data= prostate)
mod1_aic = step(mod_1, direction ="backward")
mod1_aic$call


#From what I am observing the best model would be m(formula = lpsa ~ lcavol + lweight + age + lbph + svi, data = prostate)

```

**(b)** Find the model with the **best** BIC. Report the predictors that are used in the resulting model.

```{r}

mod1_bic = step(mod_1,direction = "backward", k = log(length(resid(mod_1))))
mod1_bic$call

#From what I see the preffered model is lm(formula = lpsa ~ lcavol + lweight + svi, data = prostate).

```

**(c)** Find the model with the **best** Adjusted $R^2$. Report the predictors that are used in the resulting model.

```{r}
all <- regsubsets(lpsa ~ ., data=prostate)
cbind(summary(all)$which[,-1], "Adj R-sq"=summary(all)$adjr2)
#We see a model with 7 predictors. This has a best adjusted r-squared. It's predictors are as we see lcavol, lweight, age, lbph, svi, lcp, and pgg45.
```

```{r}
pred_mod <- lm(lpsa ~ lcavol + lweight + age + lbph + svi + lcp + pgg45, data = prostate)
```

**(d)** Of the four models you just considered, some of which *may* be the same, which is the best for making predictions? Use leave-one-out-cross-validated MSE or RMSE to decide.

```{r}
rmse = function(model) {
  sqrt(mean((resid(model) / (1 - hatvalues(model))) ^ 2))
}
```

```{r}
rmse(mod_1)
rmse(mod1_aic)
rmse(mod1_bic)
rmse(pred_mod)

#We see the aic model with smallest rmse of 0.736896 will be the best for making pedictors.
```

## Exercise 2 (Goalies Data, revisited)

**(a)** Use the data found in [`goalies_cleaned.csv`](goalies_cleaned.csv) to find a "good" model for wins, `W`. Use any methods seen in class. The model should reach a `Multiple R-squared` above `0.99` using fewer than 37 parameters. Hint: you may want to look into the ability to add many interactions quickly in `R`.

```{r}
gc <- read.csv("~/Downloads/goalies_cleaned(1).csv", comment.char="#")
full_gc <- lm(W ~., data= gc)
gc_aic = step(full_gc, direction ="backward")
gc_aic$call

#BIC

gc_bic = step(full_gc,direction = "backward", k = log(length(resid(full_gc))))
gc_bic$call

#Decided Model
opt_model= lm(W~  MIN + PIM +SO * GA * SA * SV * SV_PCT, data=gc)
summary(opt_model)
summary(opt_model)$adj.r.squared

#Bigger Model

large_set <- regsubsets(W ~ ., data=gc)
cbind(summary(large_set)$which[,-1], "Adj R-squared"=summary(large_set)$adjr2)

mul_pred = lm(W ~ GA+SA+SV+SO+MIN+PIM, data = gc)

#RMSE

rmse(full_gc)
rmse(gc_aic)
rmse(gc_bic)
rmse(opt_model)
rmse(mul_pred)


```

## Exercise 3 (Body Dimensions)

For this exercise we will use the data stored in [`body.csv`](body.csv). It contains 21 body dimension measurements as well as age, weight, height, and gender on 507 individuals. The participants were primarily individuals in their twenties and thirties, with a few older men and women, all of whom proclaimed to exercise several hours a week. The variables in the dataset are:

Skeletal Measurements (all measured in cm):

- `s1` - Biacromial diameter
- `s2` - Biiliac diameter, or "pelvic breadth"
- `s3` - Bitrochanteric diameter
- `s4` - Chest depth between spine and sternum at nipple level, mid-expiration
- `s5` - Chest diameter at nipple level, mid-expiration
- `s6` - Elbow diameter, sum of two elbows
- `s7` - Wrist diameter, sum of two wrists
- `s8` - Knee diameter, sum of two knees
- `s9` - Ankle diameter, sum of two ankles

Girth Measurements (all measured in cm):

- `g1` - Shoulder girth over deltoid muscles
- `g2` - Chest girth, nipple line in males and just above breast tissue in females, mid-expiration
- `g3` - Waist girth, narrowest part of torso below the rib cage, average of contracted and relaxed position
- `g4` - Navel (or "Abdominal") girth at umbilicus and iliac crest, iliac crest as a landmark
- `g5` - Hip girth at level of bitrochanteric diameter
- `g6` - Thigh girth below gluteal fold, average of right and left girths
- `g7` - Bicep girth, flexed, average of right and left girths
- `g8` - Forearm girth, extended, palm up, average of right and left girths
- `g9` - Knee girth over patella, slightly flexed position, average of right and left girths
- `g10` - Calf maximum girth, average of right and left girths
- `g11` - Ankle minimum girth, average of right and left girths
- `g12` - Wrist minimum girth, average of right and left girths

Other Measurements:

- `Age` - in years
- `Weight` - in kg
- `Height` - in cm
- `Gender` - 0 = female, 1 = male

```{r}
body <- read.csv("~/Downloads/body.csv")

c_rmse = function(model) {
  sqrt(mean((resid(model) / (1 - hatvalues(model))) ^ 2))
}
```

**(a)** Find a good model for `Weight` using only the Skeletal variables as well as `Age`, `Height`, and `Gender`. You are not allowed to remove any data or transform the response.

```{r}
skel_mod = lm(Weight ~ s1 + s2+ s3+s4+s5+s8 + s6 + s7 + s8 + s9 + Age + Height + Gender, data = body)
skel_aic = step(skel_mod, directions= "backward", trace=0)
skel_bic= step(skel_mod, direction= "backward", k=log(length(resid(skel_mod))), trace=0)

#Both bic and aic agree 

skel_bic$call
summary(skel_bic)$adj.r.squared

skel_bic_RSS =sum(summary(skel_bic)$residuals^2)
plot(fitted(skel_bic), resid(skel_bic), col = "red", xlab = "fitted", ylab = "residual")
abline(h = 0, col = "yellow", lwd = 2)
shapiro.test(resid(skel_bic))

skel_bic= lm(Weight~Age+Height+s1+s2+s3+s4+s5+s7+s8, data=body)
summary(skel_bic)$adj.r.squared

skel_RSS =sum(summary(skel_bic)$residuals^2)
c_rmse(skel_bic)
```


**(b)** Find a good model for `Weight` using only the Girth variables as well as `Age`, `Height`, and `Gender`. You are not allowed to remove any data or transform the response.

```{r}
model_g = lm(Weight~g1+g2+g3+g4+g5+g6+g7+g8+g9+g10+g11+g12+Age+Height+Gender, data=body)
summary(model_g)$adj.r.squared

gmod_rss =sum(summary(model_g)$residuals^2)
c_rmse(model_g)

n=length(resid(model_g))
gmod_bic= step(model_g, direction= "backward", k=log(n), trace=0)
gmod_bic

summary(gmod_bic)$adj.r.squared

g_bic_rss =sum(summary(gmod_bic)$residuals^2)
c_rmse(skel_bic)
```

**(c)** Based on **(a)** and **(b)**, which set of body measurements are more useful for predicting `Weight`.

```{r}
#The second's body measurements are more useful for predicting the weights.
```

**(d)** Using all available variables, devise and justify a good model for predicting `Weight`. It should use less than 50 parameters and obtain a LOOCV RMSE less than 1.9. You are not allowed to remove any data or transform the response.

```{r}
c_rmse = function(model) {
  sqrt(mean((resid(model) / (1 - hatvalues(model))) ^ 2))
}

mall=lm(Weight~. , data=body)
(step(mall, direction="backward", trace=0))

(step(mall, direction = "backward", k=log(length(resid(mall))), trace=0))

anova((step(mall, direction="backward", trace=0)), (step(mall, direction = "backward", k=log(length(resid(mall))), trace=0)))

aic_mall= lm(Weight ~ Age+s2*s4+s5+s6*s8+g1*g2 +g3*g5 + g6*g8+g9*g10+Height*Gender, data=body)
summary(aic_mall)

c_rmse(aic_mall)

```


